---
title: "DATA 606 Data Project Proposal"
author: "Alyssa Gurkas"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

### Background
The U.S. Department of Energy prepares an annual Electric Power Report that 
includes information about energy production, sales, consumption of fossil 
fuels, environmental data, and other topics related to energy. 
Additionally, the U.S. Department of Commerce tracks state annual summary 
statistics which include GDP by state from years 1998-2023.

### Project Description 
This project will explore if states with a higher GDP are more likely to 
release CO2 into the atmosphere from energy production than states with a lower 
GDP. 


### Data Preparation

```{r load-libraries, eval=TRUE}
library(readxl)
library(tidyverse)
library(Hmisc)
library(plotly)
library(infer)
```

```{r load-datasets, echo=TRUE}
emissions <- read_excel("emission_annual.xlsx")
stategdp <- read_excel("stategdp_summary.xlsx")
```

```{r data-transformation-emissions-data, echo=TRUE}
emissions_v2 <- emissions %>% 
  filter(`Year` >=1998,
         `Producer Type` == "Total Electric Power Industry",
         `Energy Source` == "All Sources",
         `State` != "US-TOTAL")

emissions_v3 <- emissions |> 
  filter(`Year` >=1998,
         `State` != "US-TOTAL")
  
```

```{r data-transformation-stategdp, echo=TRUE}
stategdp_v2 <- stategdp %>% 
  select(-GeoFips) %>% 
  pivot_longer(!GeoName, names_to = "year", values_to = "gdp") %>% 
  filter(`GeoName` != "United States") %>% 
  mutate(`State`= state.abb[match(`GeoName`,state.name)]) %>% 
  select(-GeoName) %>% 
  drop_na()

stategdp_v2$year <- as.numeric(stategdp_v2$year)

stategdp_v3 <- stategdp_v2 %>% 
  rename("Year" = "year",
         "GDP" = "gdp")
```

```{r data-transformation-join, echo=TRUE}
state_gdp_emissions <- left_join(stategdp_v3, emissions_v2, 
                                 by = c('State','Year'))

state_gdp_emissions <- state_gdp_emissions |> 
  drop_na() |> 
  rename("CO2" = "CO2\r\n(Metric Tons)")

state_gdp_emissions_v2 <- left_join(stategdp_v3, emissions_v3, 
                                 by = c('State','Year'))

state_gdp_emissions_v2 <- state_gdp_emissions_v2 |> 
  drop_na() |> 
  rename("CO2" = "CO2\r\n(Metric Tons)")
```

```{r data-transformation-energy-prod-class, echo=TRUE}
energy_types <- unique(emissions$`Energy Source`)
producer_types <- unique(emissions$`Producer Type`)
```

### Data Exploration

```{r CO2-GDP-eda, echo=TRUE}
# data based on the summary statement does not seem evenly distributed
summary(state_gdp_emissions$CO2)
summary(state_gdp_emissions$GDP)

# histogram of CO2 
ggplot(state_gdp_emissions, aes(x=CO2, fill=State))+
  geom_histogram()+
  xlab("CO2")

# histogram of GDP
ggplot(state_gdp_emissions, aes(x=GDP, fill=State))+
  geom_histogram()+
  xlab("GDP")

# boxplot of GDP and CO2

# try to group by quartiles and put side by side 

gdp_qg <- quantile(state_gdp_emissions$GDP, probs = c(0.25,0.5,0.75,1))

state_gdp_emissions <- state_gdp_emissions |> 
  mutate(GDP_QG = case_when(GDP < 389481.00 ~ "Bottom 75%",
                            GDP >= 389481.00 ~ "Top 25%"))

# question - how do I create a box plot for each quartile? are the quartiles 
# grouped correctly? Other ideas on how to group this? 

ggplot(state_gdp_emissions, aes(x=GDP, 
                                y=CO2,
                                group= GDP_QG
                                )) +
geom_boxplot() +
xlab("GDP") +
ylab("CO2 Emissions") 
```

```{r energy-categorical-eda, echo=TRUE}
count_energy_source <- state_gdp_emissions_v2 %>% 
  group_by(`Energy Source`) %>% 
  count()

count_energy_source

count_producer_type <- state_gdp_emissions_v2 %>% 
  group_by(`Producer Type`) %>% 
  count()

count_producer_type

#include bar graph of Energy Source and Producer Type

state_gdp_emissions_v2 <- state_gdp_emissions_v2 %>% 
  filter(`Producer Type` != "Total Electric Power Industry",
         `Energy Source` != "All Sources")

ggplot(state_gdp_emissions_v2, aes(x=`Energy Source`, group = `Energy Source`))+
  geom_bar() +
  xlab("Energy Source")

ggplot(state_gdp_emissions_v2, aes(x=`Producer Type`, group = `Producer Type`))+
  geom_bar() +
  xlab("Producer Type")

```


```{r summary-statistics-plots, echo=TRUE}
describe(state_gdp_emissions)
#Note: removed DC as there is missing data 

options(scipen=999)

gdp_plot <- ggplot(state_gdp_emissions, aes(x=Year, y=GDP, color=State))+
  geom_line()+
  labs(title ="State GDP Over Time", 
       x="Year",
       y= "GDP (in billions)")+
  theme(legend.position="none")

co2_plot <- ggplot(state_gdp_emissions, aes(x=Year, y=`CO2`, color=State))+
  geom_line()+
  labs(title ="State CO2 Emissions from Energy Production Over Time", 
       x="Year",
       y= "CO2 Emissions (Metric Tons)")+
    theme(legend.position="none")
```


```{r plotly-plots, echo=TRUE}
ggplotly(gdp_plot)

ggplotly(co2_plot)

```

```{r summary-statistics, echo=TRUE}

state_gdp_emissions |> 
summarise(median_GDP = median(GDP),
            mean_GDP = mean(GDP),
            max_GDP = max(GDP),
            iqr_GDP = IQR(GDP),
            sd_GDP = sd(GDP)
  )

state_gdp_emissions |> 
summarise(median_CO2 = median(CO2),
            mean_CO2 = mean(CO2),
            max_CO2 = max(CO2),
            iqr_CO2 = IQR(CO2),
            sd_CO2 = sd(CO2)
  )

```

### Hypothesis Test

**H~0~: States with a larger GDP emit the same amount of CO2 from energy** 
**production as states with a smaller GDP.**
</br>
**H~A~: States with a larger GDP emit more CO2 from energy** 
**production than states with a smaller GDP.**

```{r t-test-gdp-CO2, echo=TRUE}
state_gdp_emissions <- as.data.frame(state_gdp_emissions)
state_gdp_emissions$GDP_QG <- as.factor(state_gdp_emissions$GDP_QG)

obs_diff <- state_gdp_emissions %>%
  specify(CO2 ~ GDP_QG) %>%
  calculate(stat = "diff in means", order = c("Top 25%", "Bottom 75%"))

null_dist <- state_gdp_emissions %>%
  specify(CO2 ~ GDP_QG) %>%
  hypothesize(null = "independence") %>%
  generate(reps = 1000, type = "permute") %>%
  calculate(stat = "diff in means", order = c("Top 25%", "Bottom 75%"))

# pairedSamplesTTest(
# formula = CO2 ~ GDP_QG + (State),
# data = state_gdp_emissions
# )
# 
# pairedSamplesTTest(
# formula = CO2 ~ GDP, # two sided formula: outcome ~ group
# data = state_gdp_emissions, # data frame
# id = "State" # name of the id variable
# )

```

```{r histo-null-hypoth, echo=TRUE}
ggplot(data = null_dist, aes(x = stat)) +
  geom_histogram()
```

The chunk below will evaluate how many `null` permutations have a difference of 
at least `obs_stat`.

```{r prop-null-diff-obs_stat, echo=TRUE}
null_dist |> 
  filter(`stat` >= obs_diff) |> 
  count()
```

There are no permutations that have a difference of at least `obs_diff`, or
43125045.

Now that the test is initialized and the null distribution formed, 
p-value will be calculated for the hypothesis test using the function
`get_p_value`.

```{r inf-weight-habit-ht-pvalue}
null_dist %>%
  get_p_value(obs_stat = obs_diff, direction = "two_sided")
```

This the standard workflow for performing hypothesis tests.

```{r get-ci, echo=TRUE}

get_confidence_interval(null_dist, 
                        level = 0.95, 
                        type = NULL, 
                        point_estimate = NULL) # how can I determine what the 
                                               # point estimate should be?

# get_confidence_interval(point_estimate = obs_mean,
#                           # at the 95% confidence level
#                           level = .95,
#                           # using the standard error
#                           type = "se")
```

Based on these results, we can conclude that the difference between the
difference in CO2 in these two groups lies somewhere between -5166621 and 
5731007. From the results of the t-test, we can conclude that that we are 95% 
certain that the mean of the two groups is somewhere between X and X. Based on 
this, I would [reject/accept] the null hypothesis.

*Question: The confidence interval is very large. What does this mean? Should I* 
*accept the null hypothesis? How can I complete the last two sentences?*

### Linear Regression 

To determine if there is a relationship between the CO2 emitted from energy
production in the respective state, and the state GDP, a scatter plot can be
used to visualize the linear relationship. 

```{r CO2-GDP-scatterplot, echo=TRUE}
scatterplot_CO2_GDP <- ggplot(state_gdp_emissions, aes(x=GDP , y=CO2, color=State)) +
  geom_point()

ggplotly(scatterplot_CO2_GDP)
```

Based on the plot produced, there is evidence that there is a relationship, but
there seems to be a few outliers. Therefore, I will continue to complete some
exploratory analysis. 


```{r CO2-GDP-cor, echo=TRUE}
state_gdp_emissions %>%
  summarise(cor(GDP, CO2))
```

```{r m1, echo=TRUE}
m1 <- lm(CO2 ~ GDP, data = state_gdp_emissions)
```

```{r m1-summary, echo=TRUE}
summary(m1)
```

```{r scatterplot-w-model, echo=TRUE}
ggplot(data = state_gdp_emissions, aes(x = GDP, y = CO2)) +
  geom_point() +
  stat_smooth(method = "lm", se = FALSE)
```

```{r residuals-plot, echo=TRUE}
ggplot(data = m1, aes(x = .fitted, y = .resid)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  xlab("Fitted values") +
  ylab("Residuals")
```

### Analysis Excluding TX, CA

```{r exl-CA-TX, echo=TRUE}
state_gdp_emissions_v3 <- state_gdp_emissions |> 
  filter(`State` != "CA",
         `State` != "TX")
```

```{r CO2-GDP-m2, echo=TRUE}
state_gdp_emissions_v3 %>%
  summarise(cor(GDP, CO2))

m2 <- lm(CO2 ~ GDP, data = state_gdp_emissions_v3)

summary(m2)

ggplot(data = state_gdp_emissions_v3, aes(x = GDP, y = CO2)) +
  geom_point() +
  stat_smooth(method = "lm", se = FALSE)

ggplot(data = m2, aes(x = .fitted, y = .resid)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  xlab("Fitted values") +
  ylab("Residuals")

car::residualPlot(model=m2)

ggplot(data = m2, aes(x = .resid)) +
  geom_histogram(bins = 30) +
  xlab("Residuals") 

ggplot(data = m2, aes(sample = .resid)) +
  stat_qq()
```

### Analysis with Log Transformation
```{r m3-log, echo=TRUE}

m3 <- lm(log(CO2) ~ GDP, data = state_gdp_emissions)

summary(m3)

ggplot(data = m3, aes(x = .fitted, y = .resid)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  xlab("Fitted values") +
  ylab("Residuals")

car::residualPlot(model=m3)

ggplot(data = m3, aes(x = .resid)) +
  geom_histogram(bins = 30) +
  xlab("Residuals") 

ggplot(data = m3, aes(sample = .resid)) +
  stat_qq()

```

###### Note On Producer Types (Business Classification)

The EIA classifies Producer Types on page 227 of the Electric Power Annual 2023
Annual Report:

1. Non-utility producers: entities that own or operate electric generating units
but are not subject to direct economic regulation of rates, such as by state
utility commissions. Non utility power producers do not have a designated 
franchised service area. In addition to entities whose primary business is the 
production and sale of electric power, entities with other primary business 
classifications can and do sell electric power. These can consist of, for 
example, manufacturing facilities and paper mills. 

2. Electric Utility (Sector 1): all regulated plants with a primary purpose of 
selling electricity in the public markets (NAICS = 22).

3. Independent Power Producers (Sector 2): All non-regulated plants with a 
primary purpose of electric power generation and a primary purpose of selling
electricity in the public markets (NAICS = 22) with no ability to cogenerate
heat and power. 

4. Electric Power, Combined Heat and Power (Sector 3): All non-regulated plants 
with a primary purpose of electric power generation and a primary purpose of 
selling electricity in the public markets (NAICS = 22) with the ability to 
cogenerate heat and power.

5. Commercial, Non-Combined Heat and Power (Sector 4): All plants with a 
commercial primary purpose with no ability to cogenerate heat and power.

6. Commercial, Combined Heat and Power (Sector 5): All plants with a commercial
primary purpose with the ability to cogenerate heat and power.

7. Industrial, Non-Combined Heat and Power (Sector 6): All plants with an 
industrial primary purpose with no ability to cogenerate heat and power.

8. Industrial, Combined Heat and Power (Sector 7): All plants with an industrial 
primary purpose with the ability to cogenerate heat and power.

**The EIA defines renewable energy within the footnote section on page 33 of the 
Electric Power Annual 2023 Annual Report:**

- Renewable includes: wood, black liquor, other wood waste, biogenic municipal 
solid waste, landfill gas, sludge waste, agriculture byproducts, other biomass, 
geothermal, solar thermal, photovoltaic energy, and wind.

The EIA classifies Energy Types within the footnote section on page 74 of the 
Electric Power Annual 2023 Annual Report:

- Coal includes: anthracite, bituminous, subbituminous, lignite, and waste coal; 
synthetic coal and refined coal; and beginning in 2011, coal-derived synthesis gas. 
Prior to 2011 coal-derived synthesis gas was included in Other Fossil Gas.

- Petroleum Liquids includes: distillate and residual fuel oils, jet fuel, 
kerosene, waste oil, and beginning in 2011, propane. Prior to 2011 propane was 
included in Other Fossil Gas.

- Petroleum Coke includes: petroleum coke-derived synthesis gas. Prior to 2011, 
petroleum coke-derived synthesis gas was included in Other Fossil Gas. Other 
Fossil Gas includes gaseous propane, blast furnace gas, other manufactured and 
waste gases derived from fossil fuels other than hydrogen.


### Research question 

**Are states with higher GDPs more likely to produce energy that emits CO2?**

### Cases 

Each case represents a states annual energy usage and GDP for the respective 
year. There are 1326 rows in the data set and eight columns:
1. Year
<br/>
2. State
<br/>
3. Producer Type
<br/>
4. Energy Source
<br/>
5. CO2 (Metric Tons)
<br/>
6. SO2 (Metric Tons)
<br/>
7. NOx (Metric Tons)
<br/>
8. GDP

### Data collection 
This project uses data from the Department of Energy's (DOE) Annual Electric 
Power Report and the U.S. Department of Commerce (DOC) State Annual Summary 
Statistics.

The Annual Electric Power Report is prepared by the Office of Energy Production,
Conversion, and Delivery (EPCD), within the U.S. Energy Information 
Administration in DOE. Data in the report is provided directly by respondents 
into DOE's information systems. For more information about the data collection 
process, please see the *Technical Notes: Data Quality and Submission* section 
of the Electric Power Annual Report: eia.gov/electricity/annual/pdf/epa.pdf, on
page 206.

The State Annual Summary Statistics provided by DOC are estimated using two data
sources: (1) wages and salaries data from the Bureau of Labor Statistics and (2) 
value-added, receipts, and payroll data from the Census Bureau's economic 
censuses. These data sources are then used to estimate the State GDP following 
the estimation methodology outlined in the DOC's Gross Domestric Product by 
State Estimation Methodology. For more information please see the methodology:
bea.gov/sites/default/files/methodologies/0417_GDP_by_State_Methodology.pdf.

### Type of study 

This is an observational study.

### Data Sources
1. [DOC/BEA's GDP and Personal Income by State](https://www.bea.gov/itable/regional-gdp-and-personal-income)
<br>
2. [DOE/EIA's Electric Power Industry Esimated Emissions by State](https://www.eia.gov/electricity/data.php)


### Response 

The response variable is state GDP and is numerical.


### Explanatory 

The explanatory variable is CO2 Emissions from Energy Production and is 
numerical.





