---
title: "DATA 606 Data Project Proposal"
author: "Alyssa Gurkas"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

### Background
The U.S. Department of Energy prepares an annual Electric Power Report that 
includes information about energy production, sales, consumption of fossil 
fuels, environmental data, and other topics related to energy. 
Additionally, the U.S. Department of Commerce tracks state annual summary 
statistics which include GDP by state from years 1998-2023.

### Project Description 
This project will explore if states with a higher GDP are more likely to 
release CO2 into the atmosphere than states with a lower GDP. 


### Data Preparation

```{r load-libraries, eval=TRUE}
library(readxl)
library(tidyverse)
library(Hmisc)
library(plotly)
```

```{r load-datasets, echo=TRUE}
emissions <- read_excel("emission_annual.xlsx")
stategdp <- read_excel("stategdp_summary.xlsx")
```

```{r data-transformation-emissions-data, echo=TRUE}
emissions_v2 <- emissions %>% 
  filter(`Year` >=1998,
         `Producer Type` == "Total Electric Power Industry",
         `Energy Source` == "All Sources",
         `State` != "US-TOTAL")
```

```{r data-transformation-stategdp, echo=TRUE}
stategdp_v2 <- stategdp %>% 
  select(-GeoFips) %>% 
  pivot_longer(!GeoName, names_to = "year", values_to = "gdp") %>% 
  filter(`GeoName` != "United States") %>% 
  mutate(`State`= state.abb[match(`GeoName`,state.name)]) %>% 
  select(-GeoName) %>% 
  drop_na()

stategdp_v2$year <- as.numeric(stategdp_v2$year)

stategdp_v3 <- stategdp_v2 %>% 
  rename("Year" = "year",
         "GDP" = "gdp")
```

```{r data-transformation-join, echo=TRUE}
state_gdp_emissions <- left_join(emissions_v2, stategdp_v3, by = c('State','Year'))

state_gdp_emissions <- state_gdp_emissions |> 
  drop_na() |> 
  rename("CO2" = "CO2\r\n(Metric Tons)")
```

### Research question 

**Are states with higher GDPs more likely to produce energy that emits CO2?**

### Cases 

Each case represents a states annual energy usage and GDP for the respective 
year. There are 1326 rows in the data set and eight columns:
1. Year
<br/>
2. State
<br/>
3. Producer Type
<br/>
4. Energy Source
<br/>
5. CO2 (Metric Tons)
<br/>
6. SO2 (Metric Tons)
<br/>
7. NOx (Metric Tons)
<br/>
8. GDP

### Data collection 
This project uses data from the Department of Energy's (DOE) Annual Electric 
Power Report and the U.S. Department of Commerce (DOC) State Annual Summary 
Statistics.

The Annual Electric Power Report is prepared by the Office of Energy Production,
Conversion, and Delivery (EPCD), within the U.S. Energy Information 
Administration in DOE. Data in the report is provided directly by respondents 
into DOE's information systems. For more information about the data collection 
process, please see the *Data Quality and Submission* section of the 
Electric Power Annual Report: eia.gov/electricity/annual/pdf/epa.pdf.

The State Annual Summary Statistics provided by DOC are estimated using two data
sources: (1) wages and salaries data from the Bureau of Labor Statistics and (2) 
value-added, receipts, and payroll data from the Census Bureau's economic 
censuses. These data sources are then used to estimate the State GDP following 
the estimation methodology outlined in the DOC's Gross Domestric Product by 
State Estimation Methodology. For more information please see the methodology:
bea.gov/sites/default/files/methodologies/0417_GDP_by_State_Methodology.pdf.

### Type of study 

This is an observational study.

### Data Sources
1. [DOC/BEA's GDP and Personal Income by State](https://www.bea.gov/itable/regional-gdp-and-personal-income)
<br>
2. [DOE/EIA's Electric Power Industry Esimated Emissions by State](https://www.eia.gov/electricity/data.php)


### Response 

The response variable is state GDP and is numerical.


### Explanatory 

The explanatory variable is CO2 Emissions from Energy Production and is 
numerical.


### Relevant summary statistics 

```{r summary-statistics-plots, echo=TRUE}
describe(state_gdp_emissions)
#Findings: DC may need to be excluded from this evaluation due to missing data.

options(scipen=999)

gdp_plot <- ggplot(state_gdp_emissions, aes(x=Year, y=GDP, color=State))+
  geom_line()+
  labs(title ="State GDP Over Time", 
       x="Year",
       y= "GDP (in billions)")+
  theme(legend.position="none")

co2_plot <- ggplot(state_gdp_emissions, aes(x=Year, y=`CO2`, color=State))+
  geom_line()+
  labs(title ="State CO2 Emissions from Energy Production Over Time", 
       x="Year",
       y= "CO2 Emissions (Metric Tons)")+
    theme(legend.position="none")
```

The graphs above are not very helpful if they are not interactive, therefore,
I am providing the plots that are interactive and if you hover you can see 
summary information. Note, the interactive plots are only accessible in HTML
files and within the Rmd. 

```{r plotly-plots, echo=TRUE}
ggplotly(gdp_plot)

ggplotly(co2_plot)

```

```{r summary-statistics, echo=TRUE}

state_gdp_emissions |> 
summarise(median_GDP = median(GDP),
            mean_GDP = mean(GDP),
            max_GDP = max(GDP),
            iqr_GDP = IQR(GDP),
            sd_GDP = sd(GDP)
  )

state_gdp_emissions |> 
summarise(median_CO2 = median(CO2),
            mean_CO2 = mean(CO2),
            max_CO2 = max(CO2),
            iqr_CO2 = IQR(CO2),
            sd_CO2 = sd(CO2)
  )

length(unique(state_gdp_emissions$State))

```


