---
title: 'Inference for numerical data'
author: "Alyssa Gurkas"
output:
  pdf_document: default
  html_document:
    includes:
      in_header: header.html
    css: ./lab.css
    highlight: pygments
    theme: cerulean
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: console
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(eval = TRUE, message = FALSE, warning = FALSE)
```

## Getting Started

### Load packages

In this lab, we will explore and visualize the data using the **tidyverse** suite of packages, and perform statistical inference using **infer**. The data can be found in the companion package for OpenIntro resources, **openintro**.

Let's load the packages.

```{r load-packages, message=FALSE}
library(tidyverse)
library(openintro)
library(infer)
```


### The data

Every two years, the Centers for Disease Control and Prevention conduct the Youth Risk Behavior Surveillance System (YRBSS) survey, where it takes data from high schoolers (9th through 12th grade), to analyze health patterns. You will work with a selected group of variables from a random sample of observations during one of the years the YRBSS was conducted.

Load the `yrbss` data set into your workspace.

```{r load-data}
data('yrbss', package='openintro')
```

There are observations on 13 different variables, some categorical and some numerical. The meaning of each variable can be found by bringing up the help file:

```{r help-nc, eval=FALSE}
?yrbss
```


1.  What are the cases in this data set? How many cases are there in our sample?
</br>
**Exercise 1 Response**
</br>
**There are 13 variables (or columns) in this data set. For the raw unproccessed data, five variables are considered qualitative and are either in the double (dbl) or integer class (`age`,`height`,`weight`,`physically_active_7d `, `strength_training_7d`). Eight variables are considered qualitative and are in the character class (`gender`, `grade`,`hispanic`, `race`, `helmet_12m`, `text_while_driving_30d`, `hours_tv_per_school_day`, `school_night_hours_sleep`). In R, even if the data is mainly numeric, such as the `grade` column, if there are values that are characters. It will consider all the values within that variable to be characters.**

Remember that you can answer this question by viewing the data in the data viewer or by using the following command:

```{r str}
glimpse(yrbss)
```

## Exploratory data analysis

You will first start with analyzing the weight of the participants in kilograms: `weight`.

Using visualization and summary statistics, describe the distribution of weights. The `summary` function can be useful.

```{r summary}
summary(yrbss$weight)
```

2.  How many observations are we missing weights from?
</br>
**Exercise 2 Response**
</br>
**Of the 13583 observations in the data set, 1004 observations are missing from the weight variable, or seven percent.**

Next, consider the possible relationship between a high schooler's weight and their physical activity. Plotting the data is a useful first step because it helps us quickly visualize trends, identify strong associations, and develop research questions.

First, let's create a new variable `physical_3plus`, which will be coded as either "yes" if they are physically active for at least 3 days a week, and "no" if not.

```{r create new var}
yrbss <- yrbss %>% 
  mutate(physical_3plus = ifelse(yrbss$physically_active_7d > 2, "Yes", "No")) 
```


3.  Make a side-by-side boxplot of `physical_3plus` and `weight`. Is there a relationship between these two variables? What did you expect and why?
</br>
**Exercise 3 Response**
</br>

```{r exercise-3-response, echo=TRUE}
ggplot(yrbss, aes(x=physical_3plus, y=weight)) +
geom_boxplot() +
xlab("Physically Active 3x Per Week") +
ylab("Weight in Kilograms") 
```

**I expected that the people who are less active would weigh more on average. However, the box plot is showing that the people who are more active, weigh slightly more, on average. **

The box plots show how the medians of the two distributions compare, but we can also compare the means of the distributions using the following to first group the data by the `physical_3plus` variable, and then calculate the mean `weight` in these groups using the `mean` function while ignoring missing values by setting the `na.rm` argument to `TRUE`.

```{r by-means}
yrbss %>%
  group_by(physical_3plus) %>%
  summarise(mean_weight = mean(weight, na.rm = TRUE))
```

There is an observed difference, but is this difference statistically significant? In order to answer this question we will conduct a hypothesis test.

## Inference

4.  Are all conditions necessary for inference satisfied? Comment on each. You can compute the group sizes with the `summarize` command above by defining a new variable with the definition `n()`.
</br>
**Exercise 4 Response**
</br>
**To apply the central limit theorem, there two conditions:**
</br>
**(1) independence: this means that the observations must be independent, and the sample is randomized.**
</br>
**(2) normality: this means that the sample size is sufficiently large such that np is greater than or equal to ten and n(1-p) is greater than ten. **
</br>
**The CDC's data methodology documentation available at https://www.cdc.gov/mmwr/pdf/rr/rr6201.pdf certifies that the data collection method is representative of its respective population. This means we can consider the independence condition is met. However, we should ensure that the normality condition is met through some analysis. To do this, we will check if the sample size is less than 30, and if there are outliers in the data.**
</br>

```{r exercise-4-response, echo=TRUE}
# checking observations in the YRBSS dataset:
yrbss %>%
  group_by(physical_3plus) %>%
  drop_na() %>% 
  summarise(n_obs = length(weight))

yrbss_v2 <- yrbss %>% 
  drop_na()

ggplot(yrbss_v2, aes(x=weight, fill=physical_3plus)) +
    geom_histogram(position="dodge") + 
    facet_wrap(vars(physical_3plus))+
    theme(
      legend.position = "none"
    )

```
</br>
**Based on the analysis and plot, we can determine that the conditions necessary for inference are satisfied. The data seems to be normally distributed, and the sample size is sufficiently large.**


5.  Write the hypotheses for testing if the average weights are different for those who exercise at least times a week and those who don't.
</br>
**Exercise 5 Response**
</br>
**H~0~ Students who are physically active 3 or more days per week weigh the same as students who are not physically active.**
</br>
**H~A~ Students who are physically active 3 or more days per week weigh less than those who don't exercise.**

Next, we will introduce a new function, `hypothesize`, that falls into the `infer` workflow. You will use this method for conducting hypothesis tests. 

But first, we need to initialize the test, which we will save as `obs_diff`.

```{r inf-weight-habit-ht-initial, tidy=FALSE, warning = FALSE}
obs_diff <- yrbss %>%
  drop_na(physical_3plus) %>%
  specify(weight ~ physical_3plus) %>%
  calculate(stat = "diff in means", order = c("yes", "no"))
```

Notice how you can use the functions `specify` and `calculate` again like you did for calculating confidence intervals. Here, though, the statistic you are searching for is the difference in means, with the order being `yes - no != 0`.

After you have initialized the test, you need to simulate the test on the null distribution, which we will save as `null`.

```{r inf-weight-habit-ht-null, tidy=FALSE, warning = FALSE}
null_dist <- yrbss %>%
  drop_na(physical_3plus) %>%
  specify(weight ~ physical_3plus) %>%
  hypothesize(null = "independence") %>%
  generate(reps = 1000, type = "permute") %>%
  calculate(stat = "diff in means", order = c("yes", "no"))
```

Here, `hypothesize` is used to set the null hypothesis as a test for independence. In one sample cases, the `null` argument can be set to "point" to test a hypothesis relative to a point estimate.

Also, note that the `type` argument within `generate` is set to `permute`, whichis the argument when generating a null distribution for a hypothesis test.

We can visualize this null distribution with the following code:

```{r}
ggplot(data = null_dist, aes(x = stat)) +
  geom_histogram()
```


6. How many of these `null` permutations have a difference of at least `obs_stat`?
</br>
**Exercise 6 Response**
</br>
**Insert your answer here**

Now that the test is initialized and the null distribution formed, you can calculate the p-value for your hypothesis test using the function `get_p_value`.

```{r inf-weight-habit-ht-pvalue}
null_dist %>%
  get_p_value(obs_stat = obs_diff, direction = "two_sided")
```

This the standard workflow for performing hypothesis tests.

7.  Construct and record a confidence interval for the difference between the weights of those who exercise at least three times a week and those who don't, and interpret this interval in context of the data.
</br>
**Exercise 7 Response**
</br>
**insert answer here**
* * *

## More Practice

8.  Calculate a 95% confidence interval for the average height in meters (`height`) and interpret it in context.
</br>
**Exercise 8 Response**
</br>
**Insert your answer here**

9.  Calculate a new confidence interval for the same parameter at the 90% confidence level. Comment on the width of this interval versus the one obtained in the previous exercise.
</br>
**Exercise 9 Response**
</br>
**Insert your answer here**


10.  Conduct a hypothesis test evaluating whether the average height is different for those who exercise at least three times a week and those who don't.
</br>
**Exercise 10 Response**
</br>
**Insert your answer here**

11.  Now, a non-inference task: Determine the number of different options there are in the dataset for the `hours_tv_per_school_day` there are.
</br>
**Exercise 11 Response**
</br>
**Insert your answer here**

12. Come up with a research question evaluating the relationship between height or weight and sleep. Formulate the question in a way that it can be answered using a hypothesis test and/or a confidence interval. Report the statistical results, and also provide an explanation in plain language. Be sure to check all assumptions, state your $\alpha$ level, and conclude in context.
</br>
**Exercise 12 Response**
</br>
**Insert your answer here**

* * *
